---
title: "Answer-Dependent Multiple-Choice Question Demo Data Set"
author: "Kenjiro Quides"
date: "`r Sys.Date()`"
output: html_document
---
# This file will walk through a demonstration dataset that is formatted in a common scantron output

## The following packages will be used throughout this demonstration
```{r error=TRUE, message=FALSE, warning=FALSE}
library(tidyr)
library(dplyr)
library(data.table)
library(ggplot2)
library(ggpubr)
library(rstatix)
```

## 0) Each line below will become a column of data that is commonly found in a scantron output
```{r error=TRUE, message=FALSE, warning=FALSE}
# Key column may be automatically or manually added. This is dependent on the equipment and/or software used to read a multiple choice exam.
# FS indicates a fundamental statement associated with a question. This can be replaced with any question category. This will likely need to be manually added to a scantron output via spreadsheet software (e.g., Microsoft Excel) or within R. 
  # Note: This workflow is only designed for one question category column.
SID <- c("12345", "23456", "34567", "45678", "56789","67890", "Key", "FS") 
Q1 <- c("A", "A", "C", "A", "A", "B", "A", "1")
Q2 <- c("A", "A", "B", "C", "C", "C", "A", "1")
Q3 <- c("B", "A", "B", "B", "D", "B", "B", "2")
Q4 <- c("C", "B", "C", "C", "D", "A", "C", "2")
Q5 <- c("B", "A", "B", "A", "B", "B", "B", "3")
Q6 <- c("D", "B", "C", "D", "D", "C", "D", "3")
Q7 <- c("B", "C", "D", "A", "B", "A", "A", "2")
Q8 <- c("A", "D", "C", "A", "C", "A", "C", "1")
Q9 <- c("A", "B", "A", "D", "B", "A", "B", "3")

#This will create the working dataset for the rest of the demo
df <- data.frame(SID, Q1, Q2, Q3, Q4, Q5, Q6, Q7, Q8, Q9) 
df
```

### We are assuming that student demographic data will be requested from an institution independent of the assessment. Thus, we will create a second dataframe to append to student performance data. 
```{r message=FALSE, warning=FALSE}
#This will create a new vector of mock demographic data
Demo <- c("coffee", "tea", "soda", "soda", "coffee", "tea", " ", " ")

#This will associate demographic data with a student ID based on the information above
df_Demo <- data.frame(SID, Demo) %>%
  #We only need data related to student ID for this new dataset so we can ignore "Key" and FS" data
  slice_head(n = nrow(df)-2)
df_Demo
```

## 1) We will now use the mock scantron output to generate individual question scores
```{r message=FALSE, warning=FALSE}
#We want to store the output in a new dataframe
df_qa<-                                       
  #df is the name of mock data scantron output
  df %>% 
  #The following lines will assign points based on the letter choice you indicate as correct within the parentheses. A new column is created and the point values are stored within that column.
  #Point values are listed as 1 here, but can be any value. 
  #If any of the criteria below are not met, the cell will be left blank
         mutate(P1 = case_when(Q1=="A" ~ 1),
                P2 = case_when(Q2=="A" ~ 1),
                P3 = case_when(Q3=="B" ~ 1),
                P4 = case_when(Q4=="C" ~ 1),
                P5 = case_when(Q5=="B" ~ 1),
                P6 = case_when(Q6=="D" ~ 1), 
                P7 = case_when(Q7=="A" ~ 1),
 #The lines below are an example of an answer-dependent grading scheme.
                P8 = case_when(Q7=="A" & Q8=="C" |
                               Q7=="B" & Q8=="A" |
                               Q7=="C" & Q8=="B" ~ 1),
                P9 = case_when(Q8=="A" & Q9=="C" |
                               Q8=="B" & Q9=="A" |
                               Q8=="C" & Q9=="B" ~ 1))
df_qa
```

## 2) Now that point values have been assigned to each question, the dataframe can be cleaned up by removing the answer choices. We will then calculate total scores.
```{r message=FALSE, warning=FALSE}
#Different exams have a different number of questions. This will automate the counting to remove the Question columns.
#First, we count to the last question
Qend<- ncol(df)
#Second, we count to the first column we want to keep
Pstart<- ncol(df)+1
#Lastly we count to the last column we want to keep
  #We multiply by 2 because there are just as many Question columns as Point columns
  #We then subtract 1 to account for the SID column
  #There is also an option to account for removal of a question from the exam. It is not used here.
Pend<- Qend*2-1-(0)

# The FS cells were originally associated with the Question columns, but we now need them in the Point columns. This will copy FS to the Points columns (2:1+nQ; 2+nQ:2+2nQ)
df_qa[nrow(df_qa), Pstart:Pend] = as.list(as.numeric(df[nrow(df_qa), 2:Qend])) 

#We can now remove the Question columns and store in a new dataframe.
df_a<-
  df_qa%>% select(c(1, Pstart:Pend))
df_a

#To calculate the total score, we can call on columns that contain "P" for a summed score
#This can still be done before removing the Question columns as well.
df_a <- df_a %>%
  mutate(total = rowSums(select(., contains("P")),na.rm = TRUE)) 
df_a
```

### Many scantron outputs will also provide an answer choice distribution. This will mimic that output.
```{r message=FALSE, warning=FALSE}
###The scantron output is first transposed
df_t <- transpose(df[,-1])

#Transposing loses some information. This will restore lost information
colnames(df_t) <- df$SID

#We are making sure the FS column is interpreted numerically. This will be important during organization later on.
df_t$FS <- sapply(df_t$FS, as.numeric)

#The next steps require a formulat to calcuate proportion of different answer choices

#The first step calculates the total number of students.
#It is equal to last column to count, -2 to remove Key and FS columns
totalN <- ncol(df_t)-2

#This is the function to calculate proportion of answer choices that match the Key column
  #indicated by +1
Prop<- function(x){
  (sum(df_t[x, 1:totalN] %in% df_t[x, totalN+1]))/totalN}

#A new column is created using the function above
df_t$Pcorrect <- sapply(1:nrow(df_t), Prop)
df_t

#Additional columns are created that count the proportion of answer choices A-E typically offered on an MCQ
  #The correct answer choice should have the same proportion as the Pcorrect column above
df_t$prop.A <- apply(df_t[1:totalN], 1, function(x) length(which(x=="A"))/totalN)
df_t$prop.B <- apply(df_t[1:totalN], 1, function(x) length(which(x=="B"))/totalN)
df_t$prop.C <- apply(df_t[1:totalN], 1, function(x) length(which(x=="C"))/totalN)
df_t$prop.D <- apply(df_t[1:totalN], 1, function(x) length(which(x=="D"))/totalN)
df_t$prop.E <- apply(df_t[1:totalN], 1, function(x) length(which(x=="E"))/totalN)

#This will trim and store the answer choice distributions in a new dataframe.
df_dist <- df_t%>%
  #totalN represent total number of students from above. We can exclude all those columns
  select(-c(1:totalN))
df_dist
```

### This will calculate mean score per question instead of proportion correct.
```{r message=FALSE, warning=FALSE}
#Generating transposed answer choice dataframe to join with df_dist. This calculates mean score per question when partial credit or answer-dependent questions are used.
  #Answer-dependent questions may have multiple pathways to full credit.
  #The above distribution is limited to matching correctness to a "Key" vector 

#The point output is first transposed
df_at <- transpose(df_a[,-1]) #ignores SID for transpose 
colnames(df_at) <- df_a$SID

#This is just a count to remove the key row. -1 for Key
df_at<- slice_head(df_at, n=nrow(df_at)-1) 
colnames(df_at) <- df_a$SID

#ignores SID as first column
rownames(df_at) <- colnames(df_a[2:(ncol(df_a)-1)]) 

df_at<- df_at %>%
  #new column populated with question ID, allows for joining if needed
  mutate(KEYQ = colnames(df_a[2:(ncol(df_a)-1)])) 

df_at <- df_at %>%
  rowwise() %>%
  #Cells were originally left blank if no points were earned. This will replace blanks with 0
  mutate(across(where(is.numeric), ~replace(., is.na(.), 0))) %>%
  # -3 removes Key, FS, and KEYQ columns
  mutate(mean = mean(c_across(1:(ncol(df_at)-3)))) %>% 
  # -4 removes the new "mean" column as well
  mutate(sum = sum(c_across(1:(ncol(df_at)-4)))) 
df_at

#This will create the new dataframe with question distribution outcomes 
df_dist <- bind_cols(df_dist, df_at$mean)
df_dist <- df_dist %>%
  rename(mean = ncol(df_dist))
df_dist
```

## 3) The following step will allow for aggregate analysis based on a Question category (e.g., FS or Bloom's)
```{r message=FALSE, warning=FALSE}
#Reset df_at for aggregate analysis
df_at <- transpose(df_a[,-1]) 
colnames(df_at) <- df_a$SID

df_at <- df_at %>%
  #Cells were originally left blank if no points were earned. This will replace blanks with 0
  replace(is.na(.), 0)

#starting at col2 skips SID col1; adds "Question" column 
df_at$Question <- colnames(df_a[2:ncol(df_a)]) 

#Note: Replace FS with Bloom as needed
df_atg<-gather(df_at, key = "SID", value = "total", -Key, -FS, -Question)

#This is where you aggregate student scores Question Category (e.g., FS, Bloom) . Recommend changing the df name based on variable you are collapsing
FS_score<-df_atg %>%   
  group_by(FS, SID) %>%
  #creates a sum column based on points assigned
  summarise(FS_score=sum(total,na.rm=TRUE))
FS_score

FS_poss<- df_atg %>%
  group_by(FS, SID) %>%
  #creates a sum column based on total points possible (indicated by Key points) 
  summarise(FS_points_poss=sum(Key))
FS_poss

#Joins newly created dataframes
df_FS <- left_join(FS_score, FS_poss)

#Calculates a proportional correct column
df_FS<- df_FS %>%
  mutate(FS_score2 = FS_score/FS_points_poss)

#removes summed FS values. There are three FS in the example, so any values above this are sums
df_FS<- df_FS %>%
  filter(FS < 4) 
df_FS
```

## 4) Demographic data can also be incorporated for further assessment
```{r message=FALSE, warning=FALSE}
#Incorporating student demographic data
df_FS_Demo <- inner_join(df_FS, df_Demo, by = "SID") 

#It may be advantageous to treat FS as a character, not numeric
df_FS_Demo$FS <- sapply(df_FS_Demo$FS, as.character)
#This will ensure FS are displayed in order
df_FS_Demo$FS <- factor(df_FS_Demo$FS, levels = c("1", "2", "3"))

#You can add a more meaningful name to each FS or Question category
df_FS_Demo <- df_FS_Demo %>%
  mutate(Theme = case_when(FS == "1"  ~ "one",
                           FS == "2"  ~ "two",
                           FS == "3"  ~ "three"))

#This will ensure FS are displayed in order
df_FS_Demo$Theme <- factor(df_FS_Demo$Theme, levels = c("one",
                                                        "two",
                                                        "three"))
df_FS_Demo
```

## 5) Data Visualization with Question Category and Demographic incorporated 
```{r message=FALSE, warning=FALSE}
#Graphing the results of each Demographic
p <- ggplot(df_FS_Demo, aes(x= Demo, y=FS_score2, color = Demo))+
  theme_classic2()+   
  geom_boxplot()+
  scale_y_continuous(limits = c(0,1.01))+
#optional line breaks, useful with 5+ categories
  #geom_vline(xintercept = c(0.5+(1:4)), linetype="dotted", color = "grey", size=0.5)+ 
#optional mean score
  #geom_hline(yintercept = mean(df_FS_Demo$FS_score2), linetype="dotted", color = "grey", size=0.3)+ 
  ggtitle("Demographic Analysis")+
  xlab("Demo")+
  ylab("FS Proportion Correct")+
  theme(legend.position = "bottom")+
#Creates a separate panel for each FS with a meaningful name
  facet_wrap(~Theme)
p
```

###  Addition of statistical comparison to data visualizations
```{r message=FALSE, warning=FALSE}
###Addition of statistical comparisons on graph
df<- df_FS_Demo
stat.test <- df %>%
  #Check grouping (Theme or Demo)
  group_by(Theme)%>%
  #Check comparison (Theme or Demo)
  wilcox_test(FS_score2 ~ Demo) %>%              
  adjust_pvalue(method = "bonferroni") %>%
  add_significance() 
stat.test 

stat.test <- stat.test %>% add_xy_position(x = "Theme", dodge = 0.8)

p + stat_pvalue_manual(stat.test, label = "p.adj.signif", t.length = 0.01,
                       hide.ns = TRUE, y.position = 0.85)+
  scale_y_continuous(expand = expansion(mult = c(0.01, 0.1)))
```